A Guide to Pre-Processing High-Throughput Animal Tracking Data

### 2021-02-04 docs/ms_atlas_preproc_main_text.tex
edit figure 2 caption

    Package functions are prefixed `\texttt{atl\_}', and are based on the \texttt{data.table} package (Dowle and Srinivasan 2020).
    Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
    Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area (see \textsc{Spatio-Temporal Filtering}).
    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
    \item Users should then reduce small-scale location errors by applying a median smooth (see \textsc{Smoothing and Thinning Data}).
    \item Users who need uniformly thinned data can either aggregate or resample it (see \textsc{Smoothing and Thinning Data}).
    At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.
    \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).
    % Residence patches are spatio-temporally independent clusters of positions that represent prolonged residence of a tracked individual in an area, and which can indicate aspects of space use such resource exploitation \citep{barraquand2008, bijleveld2016, oudman2018}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
    \caption{\textbf{(a)} A velocity-autocorrelated movement track simulated for 5,000 positions (black line) using the \texttt{smoove} package (Gurarie et al. 2017). Three kinds of errors have been artificially added: (1) each position (grey points) is offset from the canonical track with the addition of normally distributed small-scale error, (2) large-scale error has been added to 0.5\% of positions, and (3) 300 positions (indices 500 -- 800) have been displaced to the top-right of the track to simulate a gross distortion that affects a continuous subset of the track. 
    The goal of pre-processing such datasets is to get the estimated positions (grey points) to match the canonical track (black line) as closely as possible.  
    Tracks can be quickly filtered by spatial bounds ($0.5 \leq Y \leq 1.0$; dashed grey lines) using the \texttt{atlastools} function  \texttt{atl\_filter\_bounds}.
    Setting the function argument \texttt{remove\_inside = FALSE} retains positions within user supplied bounds (dashed grey lines; green points), and excludes those outside (grey points).
    Reducing large-scale outliers in a movement track. The goal is to remove positions far away from the canonical track (black line), and retain those with only reasonable displacements from the canonical track (green points). 
    \textbf{(b)} Location error may affect single observations resulting in point outliers or `spikes' (grey circles), but it may also affect continuous subsets of a track, called a `prolonged spike' (purple triangles). 
    The former may be targeted by filtering on appropriate covariates such as speed and turning angle using the \texttt{atlastools} function \texttt{atl\_filter\_covariates}.
    Here, we filter out positions with incoming and outgoing speeds $\geq$ the 90\textsuperscript{th} percentile (0.025) and turning angle $\geq$ the 10\textsuperscript{th} percentile (35$^{\circ}$).
    Howevever, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
    The \texttt{atlastools} function \texttt{atl\_remove\_reflections} to remove prolonged spikes (grey circles) in tracking data is an example of targeting track subsets with a common distortion (here, a geometric translation).
    While this method filters out the prolonged spike to return only positions with reasonable displacement from the canonical track (green points) in this example, conceptualising and implementing such algorithms is difficult. 
    Users are cautioned to frequently check this and similar methods' results.
    }
    \label{fig:figure_simulated_data}
\end{figure}

>>>>

    Package functions are prefixed `\texttt{atl\_}', and are based on the \texttt{data.table} package (Dowle and Srinivasan 2020).
    Certain functions modify the input data "in place", and users must work on a copy of their data to preserve both the original and cleaned data (details in the Supplementary Material).
    Since high-throughput tracking systems typically cover an area of a few hundred square kilometres, \texttt{atlastools} users must convert their geographic coordinates to metres as this is assumed by all functions.
    \item Users can use simple spatio-temporal range filters to select positions within a certain time or area (see \textsc{Spatio-Temporal Filtering}).
    \item Next, users should reduce gross location errors by removing unreliable positions which may be identified by a system-specific error measure, or by the plausibility of associated movement metrics (see \textsc{Filtering to Reduce Location Errors}).
    \item Users should then reduce small-scale location errors by applying a median smooth (see \textsc{Smoothing and Thinning Data}).
    \item Users who need uniformly thinned data can either aggregate or resample it (see \textsc{Smoothing and Thinning Data}).
    At this stage, the data are ready for a number of popular statistical treatments such as Hidden Markov Model-based classification \citep{michelot2016,langrock2012}.
    \item Finally, users wishing to study prolonged animal residence in an area can classify their data into residence patches based on the movement ecology of their study system, after filtering out non-stationary positions (see \textsc{Synthesising Movement Tracks into Residence Patches}).

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{figures/fig_02_filtering_data.png}
    \caption{\textbf{(a)} Simulated movement data (black line), with three kinds of artificially added errors: (1) normally distributed small-scale error on each position, (2) large-scale error added to 0.5\% of positions, and (3) 300 consecutive positions displaced to simulate a gross distortion affecting a continuous subset of the track.  
    Tracks can be quickly filtered by spatial bounds (dashed grey lines) to exclude broad regions.
    \textbf{(b)} 
    Location error may affect single observations resulting in point outliers or `spikes' (grey crosses), or continuous subsets of a track, called a `prolonged spike' (black crosses, top right), and both represent unrealistic movement.
    While spikes can be removed by filtering out positions with high incoming and outgoing speeds and turning angle, prolonged spikes cannot be effectively corrected by targeting each coordinate pair in isolation.
    These and other distortions that affect continuous track segments can be removed by conceptualising algorithms that find the bounds of the distortion and remove points in between them.
    Users should frequently check the outputs of such algorithms to avoid rejecting valid data.
    }
    \label{fig:figure_simulated_data}
\end{figure}
